Estimating duration in the presence of misclassification

### 2024-07-29 ATACCC.tex
Rob p74 comment

Biological plausibility requires constraining $\vec{\phi}_i$ to positive real values.
Each $\vec{\phi}_i$ is modelled as a log-normal random effects by using the hierarchical prior $\vec{\phi}_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
$\vec{\mu}_{\phi,g} = [\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g}]^T$ is a vector of the population means of the log-parameters.
As the random effects have a log-normal, rather than normal, distribution $\exp(\vec{\mu}_{\phi,g})$ is the population median of the parameter value but not the mean.
The covariance matrix $\matr{\Sigma}_\phi$ controls the variability of the population and allows for the parameters to be correlated.
For example, an individual with a high peak viral load may also clear the virus slowly because both have an underlying cause of a weak immune system.
To improve sampling efficiency, the multivariate version of the non-centred parameterization~\autocite{papaspiliopoulosGeneral,stanReparameterization} is used.
This can be expressed as follows.
\begin{align}
  \vec{\eta}_i &\dist \N(\vec{0}, \matr{C}_\phi) \\
  \log \vec{\phi}_i &= \vec{\mu}_\phi + \vec{\sigma}_\phi \circ \vec{\eta}_i
\end{align}
where $\vec{\sigma}_\phi$ is a vector of the standard deviations of the parameters, and $\matr{C}_\phi$ is a correlation matrix.
These are defined such that, by construction, $(\matr{\Sigma}_\phi)_{ij} = (\vec{\sigma}_\phi)_i (\vec{\sigma}_\phi)_j (\matr{C}_\phi)_{ij}$.
% Correlation matrices have diagonal elements all 1, therefore this reparameterization has the same number of parameters as the original model.

@@

From \cref{ATACCC:eq:di}, we can see that $d_i$ depends only on $i$'s random effects vector $\vec{\phi}_i$.
Therefore, the distribution of $d_i$ across the population is determined by $\vec{\phi}_i$'s mean, $\vec{\mu}_\phi$, and covariance matrix, $\matr{\Sigma}_\phi$.
% Hence, the uncertainty in the posterior distribution of the discretised duration cumulative distribution function for group $g$, $F_{D,g}(d) = \prob(D_i \leq d + 0.5)$ originate from the uncertainty in $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.

For a known $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$, $F_{D,g}$ is found as:
\begin{align}
  F_{D,g}(d)
  &= \prob(D_i \leq d + 0.5) \\
  &= \E_{D_i}(\I(D_i \leq d + 0.5))
  \label{ATACCC:eq:FDintegral}
\end{align}
where $\I$ is the indicator function.
The integral in \cref{ATACCC:eq:FDintegral} can be approximated using Monte Carlo integration (see \cref{ATACCC:alg:MonteCarlo}).
Equivalent forms of $F_{D,g}$, such as the survival and hazard functions, can be calculated using the relations in \cref{E-inc-prev:sec:definitions}.

\begin{algorithm}
\For{$i = 1, \dots, N$}{
  \For{$j = 1, \dots, M$}{
    Sample an individual-level random effects vector $\vec{\phi}_{i,j} \dist \LN(\vec{\mu}_{\phi,g}, \matr{\Sigma}_\phi)$ \;
    Calculate $d_{i,j}$ using \cref{ATACCC:eq:di} \;
  }
  \For{$d = 1, \dots, \dmax$}{
    Set $F_{D,g}^{(i)}(d) = \frac{1}{M} \sum_{j=1}^M \I(d_{i,j} \leq d + 0.5)$ \;
  }
}
  \caption{%
  Algorithm to calculate the Monte Carlo integral in \cref{ATACCC:eq:FDintegral}, producing the posterior distribution of $F_{D,g}(d)$.
  The posterior distribution is represented as a sample $F_{D,g}^{(1)}, \dots, F_{D,g}^{(N)}$.
  $\vec{\mu}_\phi^{(i)}$ and $\matr{\Sigma}_\phi^{(i)}$ denote the $i$th posterior sample of $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$ respectively, taken from the MCMC output which produced $N$ posterior samples.
  $M$ is the number of individuals to draw, I found $M =\numprint{100000}$ produced stable estimates of the tail of $F_{D,g}(d)$ if $\dmax = 45$.
 }
 \label{ATACCC:alg:MonteCarlo}
\end{algorithm}

Applying \cref{ATACCC:alg:MonteCarlo} produces a sample of the posterior distribution of $F_{D,g}(d)$.
One sample is produced for each iteration of the MCMC output used to estimate $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.

>>>>

Biological plausibility requires constraining $\vec{\phi}_i$ to positive real values.
Each $\vec{\phi}_i$ is modelled as a log-normal random effects by using the hierarchical prior $\vec{\phi}_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
$\vec{\mu}_{\phi,g} = [\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g}]^T$ is a vector of the population means of the log-parameters.
As the random effects have a log-normal, rather than normal, distribution $\exp(\vec{\mu}_{\phi,g})$ is the population median of the parameter value but not the mean.
The covariance matrix $\matr{\Sigma}_\phi$ controls the variability of the population and allows for the parameters to be correlated.
For example, an individual with a high peak viral load may also clear the virus slowly because both have an underlying cause of a weak immune system.
To improve sampling efficiency, the multivariate version of the non-centred parameterization~\autocite{papaspiliopoulosGeneral,stanReparameterization} is used.
This can be expressed as follows.
\begin{align}
  \vec{\eta}_i &\dist \N(\vec{0}, \matr{C}_\phi) \\
  \log \vec{\phi}_i &= \vec{\mu}_{\phi,g(i)} + \vec{\sigma}_\phi \circ \vec{\eta}_i
\end{align}
where $\vec{\sigma}_\phi$ is a vector of the standard deviations of the parameters, and $\matr{C}_\phi$ is a correlation matrix.
These are defined such that, by construction, $(\matr{\Sigma}_\phi)_{ij} = (\vec{\sigma}_\phi)_i (\vec{\sigma}_\phi)_j (\matr{C}_\phi)_{ij}$.
% Correlation matrices have diagonal elements all 1, therefore this reparameterization has the same number of parameters as the original model.

@@

From \cref{ATACCC:eq:di}, we can see that $d_i$ depends only on $i$'s random effects vector $\vec{\phi}_i$.
Therefore, the distribution of $d_i$ across the population of group $g$ is determined by $\vec{\phi}_i$'s mean, $\vec{\mu}_{\phi,g}$, and covariance matrix, $\matr{\Sigma}_\phi$.
% Hence, the uncertainty in the posterior distribution of the discretised duration cumulative distribution function for group $g$, $F_{D,g}(d) = \prob(D_i \leq d + 0.5)$ originate from the uncertainty in $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.

For a known $\vec{\mu}_{\phi,g}$ and $\matr{\Sigma}_\phi$, $F_{D,g}$ is found as:
\begin{align}
  F_{D,g}(d)
  &= \prob(D_i \leq d + 0.5) \\
  &= \E_{D_i}(\I(D_i \leq d + 0.5))
  \label{ATACCC:eq:FDintegral}
\end{align}
where $\I$ is the indicator function.
The integral in \cref{ATACCC:eq:FDintegral} can be approximated using Monte Carlo integration (see \cref{ATACCC:alg:MonteCarlo}).
Equivalent forms of $F_{D,g}$, such as the survival and hazard functions, can be calculated using the relations in \cref{E-inc-prev:sec:definitions}.

\begin{algorithm}
\For{$i = 1, \dots, N$}{
  \For{$j = 1, \dots, M$}{
    Sample an individual-level random effects vector $\vec{\phi}_{i,j} \dist \LN(\vec{\mu}_{\phi,g}, \matr{\Sigma}_\phi)$ \;
    Calculate $d_{i,j}$ using \cref{ATACCC:eq:di} \;
  }
  \For{$d = 1, \dots, \dmax$}{
    Set $F_{D,g}^{(i)}(d) = \frac{1}{M} \sum_{j=1}^M \I(d_{i,j} \leq d + 0.5)$ \;
  }
}
  \caption{%
  Algorithm to calculate the Monte Carlo integral in \cref{ATACCC:eq:FDintegral}, producing the posterior distribution of $F_{D,g}(d)$.
  The posterior distribution is represented as a sample $F_{D,g}^{(1)}, \dots, F_{D,g}^{(N)}$.
  $\vec{\mu}_{\phi,g}^{(i)}$ and $\matr{\Sigma}_\phi^{(i)}$ denote the $i$th posterior sample of $\vec{\mu}_{\phi,g}$ and $\matr{\Sigma}_\phi$ respectively, taken from the MCMC output which produced $N$ posterior samples.
  $M$ is the number of individuals to draw, I found $M =\numprint{100000}$ produced stable estimates of the tail of $F_{D,g}(d)$ if $\dmax = 45$.
 }
 \label{ATACCC:alg:MonteCarlo}
\end{algorithm}

Applying \cref{ATACCC:alg:MonteCarlo} produces a sample of the posterior distribution of $F_{D,g}(d)$.
One sample is produced for each iteration of the MCMC output used to estimate $\vec{\mu}_{\phi,g}$ and $\matr{\Sigma}_\phi$.
